# Run charts {#runcharts}

Run charts are designed to show a metric of interest over time. They do not rely on parametric statistical theory, so they cannot distinguish between common cause and special cause variation. Control charts can be more powerful when properly constructed, but run charts are easier to implement where statistical knowledge is limited and still provide practical monitoring and useful insights into the process.

Run charts typically employ the median for the reference line. Run charts help you determine whether there are unusual runs in the data, which suggest non-random variation. They tend to be better than control charts in trying to detect moderate (~1.5$\sigma$) changes in process than using the control charts' typical 3$\sigma$ limits rule alone. In other words, a run chart can be more useful than a control chart when trying to detect improvement while that improvement work is still going on.

## Interpreting run charts

There are two basic "tests" for run charts (an astronomical data point or looking for cycles aren't tests *per se*):  

- *Process shift:* A non-random run is a set of $log_2(n) + 3$ consecutive data points (rounded to the nearest integer) that are all above or all below the median line, where *n* is the number of points that do *not* fall directly on the median line. For example, if there are 34 points and 2 fall on the median, then $n = 32$ observations. Thus in this case, the longest run should be no more than 8 points.  

- *Number of crossings:* Too many or too few median line crossings suggest a pattern inconsistent with natural variation. You can use the binomial distribution (`qbinom(0.05, n-1, 0.50)` in R for a 5% false positive rate and expected proportion 50% of values on each side of the median) or a table (e.g., [Table 1 in Anhøj & Olesen 2014](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0113825#pone-0113825-t001)) to find the minimum number of expected crossings. Using the same data as in the example introduced in [Chapter 1](#where), we would expect the time series to cross the median entirely at least 11 times (points on the median that return to the same side are not crossings).  

```{r qichart, fig.width=8, fig.height=3.5}
# The qic function in the qicharts package creates easy run and control charts
# The runvals option shows the test results on the plot
qicharts::qic(point.signal, runvals=TRUE, ylab="Value", main="")
```


## Run charts with a trend {#runtrend}

When you first notice a trend like that seen above, you can generally wait until the process has settled to a new, stable mean and reset the central line accordingly. For a sustained or continuous trend, you can difference the data (create a new dataset by subtracting the value at time *t* from the value at time *t+1*) to remove the trend or use regression residuals to show deviations from the trend. However, either approach can make the run chart harder to interpret. Perhaps a better idea is use quantile regression to obtain the median line, which allows you to keep the data on the original scale.  

```{r quantreg, fig.height=3}
# Generate fake process data with an upward trend

# Do we need to repeat this since we made it before?
# Might serve as a good reminder, or maybe just delete the first
# instance of it to save space.

set.seed(81)
n = 36
x = seq(1:n) 
mb = data.frame(Subgroup = seq(as.Date("2006-01-01"), by = "month", length.out = n), 
                Value = 10000 + (seq(1:n) * 1.25) + (rnorm(n, 0, 5)))


# Plot with quantile regression for median line
ggplot(mb, aes(Subgroup, Value)) + 
  xlab("Subgroup") + 
  ylab("Value") + 
  geom_line(color="gray70") +
  geom_point() +
  geom_quantile(quantiles = 0.5)
```


**Process shift "test"**: $log_2(n) + 3$ where $n$ is 33 points that do not touch the median is `r round(log2(33) + 3, 2)`, so there should be no more than 8 points in any given run. The longest run in this chart is 5 points.    

**Crossings "test"**: `qbinom(0.05, 33, 0.50)` is `r qbinom(0.05, 33, 0.50)`, the minimum number of crossings we'd expect. There are 13 crossings in this run chart, as points that fall on the median line and return to the same side are not considered crossings.     

Both tests suggest there is no non-random variation in this process.  


