# Control charts for numeric, normally-distributed data {#numeric}


| If your data involve... | use a/an ... | based on the ... distribution. | 
| --------------------------------- | --------- | ----------------------------- | 
| Individual points | *I* chart | normal | 
| Subgroup average | $\bar{x}$ and *s* chart | normal |
| Exponentially weighted moving average | EWMA chart | normal |
| Cumulative sum | CUSUM chart | normal |


- For continuous data, the definition of the control limits will depend on your question and the data at hand. To detect small shifts in the mean quickly, an EWMA is probably best, while to understand natural variation and try to detect special cause variation, an $\bar{x}$ and *s* chart will be more useful.

- In the rare cases you may need an individual chart, do *not* use 3$\sigma$ for the control limits; you must use 2.66$MR_{bar}$ instead to ensure the limits are presented correctly.  

- Note: EWMA and CUSUM charts aren't "standard" control charts in that the only guideline for detecting special cause variation is a point outside the limits. So while they can't detect special cause variation like control charts, they *can* detect shifts in mean with fewer points than a standard control chart. ***WHICH TO PREFER AND WHY?***    

<br/>  

# INTERPRET MR and s charts since they don't follow usual value axis interp

## *I-MR* chart

When you have a single measurement per subgroup, the *I-MR* combination chart is appropriate. They should always be used together.  

**Mean($\bar{x}$):** &nbsp;&nbsp; $\bar{x} = \frac{\sum_{x_{i}}}{n}$

**Control limits for normal data (*I*):** 2.66$MR_{bar}$  
&nbsp;&nbsp;&nbsp;&nbsp; *where*  
&nbsp;&nbsp;&nbsp;&nbsp; $MR_{bar}$ = average moving range of *x*s, excluding those > 3.27$MR_{bar}$   

<br/>  
    
*Lab results turnaround time*
    
```{r imrex, fig.height = 3.5}
# Generate sample data
arrival = cumsum(rexp(24, 1/10))
process = rnorm(24, 5)
exit = matrix( , length(arrival))
exit[1] = arrival[1] + process[1]

for (i in 1:length(arrival)) {
    exit[i] = max(arrival[i], exit[i - 1]) + process[i]
}

# Calculate control chart inputs
subgroup.i = seq(1, length(exit))
subgroup.mr = seq(1, length(exit) - 1)

point.i = exit - arrival
point.mr = matrix(, length(point.i) - 1)
for (i in 1:length(point.i) - 1) {
    point.mr[i] = abs(point.i[i + 1] - point.i[i])
}

mean.i = mean(point.i)
mean.mr = mean(point.mr)
sigma.i = rep(mean.mr, length(subgroup.i))
sigma.mr = rep(mean.mr, length(subgroup.mr))

# Plot MR chart
spc.plot(subgroup.mr, point.mr, mean.mr, sigma.mr, k = 3.27, 
         lcl.show = FALSE, band.show = FALSE, 
         label.x = "Test number", label.y = "Turnaround time (moving range)")

# Plot I chart
spc.plot(subgroup.i, point.i, mean.i, sigma.i, k = 2.66,
         lcl.min = 0, band.show = FALSE, 
         label.x = "Test number", label.y = "Turnaround time")

```

<br/>

## $\bar{x}$ and *s* chart

When you have a sample or multiple measurements per subgroup, the $\bar{x}$ and *s* chart combination is the appropriate choice. As with the *I-MR* chart, they should always be used together.  

Control limits (3&sigma;) are calculated as follows:  
 		  
**Variable averages ($\bar{x}$):** &nbsp;&nbsp; $3\frac{\bar{s}}{\sqrt{n_i}}$
 		  
**Variable standard deviation (*s*):** &nbsp;&nbsp; $3\bar{s}\sqrt{1-c_4^2}$  
&nbsp;&nbsp;&nbsp;&nbsp; *where* $c_4 = \sqrt{\frac{2}{n-1}}\frac{\Gamma(\frac{n}{2})}{\Gamma(\frac{n-1}{2})}$  

<br/>  

*Patient wait times*   

``` {r xex, fig.height = 3.5}
# Generate sample data
set.seed(777)
waits = c(rnorm(1700, 30, 5), rnorm(650, 29.5, 5))
months = strftime(sort(as.Date('2013-10-01') + sample(0:729, 
    length(waits), TRUE)), "%Y-%m-01")
sample.n = as.numeric(table(months))
dfw = data.frame(months, waits)

# Calculate control chart inputs
subgroup.x = as.Date(unique(months))
subgroup.s = subgroup.x
point.x = aggregate(dfw$waits, by = list(months), FUN = mean, na.rm = TRUE)$x
point.s = aggregate(dfw$waits, by = list(months), FUN = sd, na.rm = TRUE)$x
mean.x = mean(waits)
mean.s = sqrt(sum((sample.n - 1) * point.s ^ 2) / 
        (sum(sample.n) - length(sample.n)))
sigma.x = mean.s / sqrt(sample.n)
c4 = sqrt(2 / (sample.n - 1)) * gamma(sample.n / 2) / 
  gamma((sample.n - 1) / 2)
sigma.s = mean.s * sqrt(1 - c4 ^ 2)

# Plot s chart
spc.plot(subgroup.s, point.s, mean.s, sigma.s, k = 3,
         label.x = "Month", label.y = "Wait times standard deviation (s)")

# Plot xbar chart
spc.plot(subgroup.x, point.x, mean.x, sigma.x, k = 3,
         label.x = "Month", label.y = "Wait times average (x)")
```

 
<br/>  

## EWMA chart

**Control limits for exponentially weighted moving average (EWMA):**  $3\frac{\bar{s}}{\sqrt{n_i}}\sqrt{\frac{\lambda}{2-\lambda}[1 - (1 - \lambda)^{2i}]}$   
&nbsp;&nbsp;&nbsp;&nbsp; *where* $\lambda$ is a weight that determines the influence of past observations. If unsure choose $\lambda = 0.2$, but $0.05 \leq \lambda \leq 0.3$ is acceptable (where larger values give stronger weights to past observations).

<br/>  

*Patient wait times (continued)*  
 
``` {r ewmaex, fig.height = 3.5}
# Use generated sample data from xbar chart example
 
# Calculate control chart inputs
subgroup.z = subgroup.x
lambda = 0.2
point.z = matrix( , length(point.x))
point.z[1] = mean.x
for (i in 2:length(point.z)) {
     point.z[i] = lambda * point.x[i] + (1 - lambda) * point.z[i-1]
     }
mean.z = mean.x
sigma.z = (mean.s / sqrt(sample.n)) * 
     sqrt(lambda/(2-lambda) * (1 - (1-lambda)^(seq(1:length(point.z)))))
 
# Plot EWMA chart
spc.plot(subgroup.z, point.z, mean.z, sigma.z, k = 3, band.show = FALSE, 
         rule.show = FALSE, label.x = "Month", 
         label.y = "Wait times moving average")
```

<br/>  

## CUSUM

Lower and upper cumulative sums are calculated as follows:

$S_{l,i} = -\max{[0, -z_i -k + S_{l,i-1}]},$  
$S_{h,i} = \max{[0, z_i -k + S_{h,i-1}]}$  
&nbsp;&nbsp;&nbsp;&nbsp; *where* $z_i$ is the standardized normal score for subgroup $i$ and $0.5 \leq k \leq 1$ is a slack value.   

It is common to choose "decision limits" of $\pm 4$ or $\pm 5$.  

<br/>  


``` {r cusumex, fig.height = 3.5}
# Use generated sample data from xbar chart example

# Calculate control chart inputs
subgroup.cusum = subgroup.x
slack = 0.5

zscore = (point.x - mean.x)/sigma.x
point.cusuml = matrix(, length(zscore))
point.cusuml[1] = -max(0, -zscore[1] - slack)

for (i in 2:length(point.cusuml)) {
    point.cusuml[i] = -max(0, -zscore[i] - slack - point.cusuml[i-1])
}

point.cusumh = matrix(, length(zscore))
point.cusumh[1] = max(0, zscore[1] - slack)

for (i in 2:length(point.cusuml)) {
    point.cusumh[i] = max(0, zscore[i] - slack - point.cusumh[i - 1])
}

mean.cusum = 0
sigma.cusum = rep(1, length(subgroup.cusum))

# Plot CUSUM chart
lower.plot = spc.plot(subgroup.cusum, point.cusuml, mean.cusum, sigma.cusum, 
        k = 5, band.show = FALSE, rule.show = FALSE, 
        label.y = "Cumulative sum")

lower.plot + geom_line(aes(y = point.cusumh), col = "royalblue3") +
    geom_point(aes(y = point.cusumh), col = "royalblue3")
```
