--- 
title: "Statistical Process Control in Healthcare with R"
author: "Dwight Barry, Brendan Bettinger, and Andrew Cooper"
date: "`r format(Sys.Date(), '%B %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: Rmadillo/spc_healthcare
description: "Using SPC methods in healthcare can be tricky. We show you how to do it correctly, using R."
---

# Overview  {#overview}

Statistical process control (SPC) was a triumph of manufacturing analytics, and its success spread across a variety of industries---most improbably, in healthcare. 

Healthcare is rarely compatible with the idea of an assembly line, but Lean thinking has taken over healthcare management around the world, and SPC methods are common tools in  Lean. This has led to much *misuse* of these methods, with predictably poor results in cases where the intent did not match the application, and bad methods lead to bad inferences. This book aims to help analysts apply SPC methods correctly, using R [@R2016]. 

```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Load libraries
library(dplyr)
library(scales)
library(lubridate)
library(forecast) 
library(ggseas)
library(qicharts)
library(bookdown)
library(knitr)
library(ggplot2)
library(ggExtra)
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'ggplot2', 'knitr', 'rmarkdown'
), 'packages.bib')
```

``` {r spccode, echo = FALSE, fig.height = 3.5}

spc.plot = function(subgroup, point, mean, sigma, k = 3,
                     ucl.show = TRUE, lcl.show = TRUE, 
                     band.show = TRUE, rule.show = TRUE,
                     ucl.max = Inf, lcl.min = -Inf,
                     label.x = "Subgroup", label.y = "Value")
{
    # Plots control chart with ggplot
    # 
    # Args:
    #   subgroup: Subgroup definition (for x-axis)
    #   point: Subgroup sample values (for y-axis)
    #   mean: Process mean value (for center line)
    #   sigma: Process variation value (for control limits)
    #   k: Specification for k-sigma limits above and below center line.
    #      Default is 3.
    #   ucl.show: Visible upper control limit? Default is true.
    #   lcl.show: Visible lower control limit? Default is true.
    #   band.show: Visible bands between 1-2 sigma limits?  Default is true.
    #   rule.show: Highlight run rule indicators in orange?  Default is true.
    #   ucl.max: Maximum feasible value for upper control limit.
    #   lcl.min: Minimum feasible value for lower control limit.
    #   label.x: Specify x-axis label.
    #   label.y: Specify y-axis label.
    
    
    df = data.frame(subgroup, point)
    df$ucl = pmin(ucl.max, mean + k*sigma)
    df$lcl = pmax(lcl.min, mean - k*sigma)
    
    warn.points = function(rule, num, den) {
        sets = mapply(seq, 1:(length(subgroup) - (den - 1)), 
                       den:length(subgroup))
        hits = apply(sets, 2, function(x) sum(rule[x])) >= num
        intersect(c(sets[,hits]), which(rule))
    }
    orange.sigma = numeric()
    
    p = ggplot(data = df, aes(x = subgroup)) +
        geom_hline(yintercept = mean, col = "gray", size = 1)
    
    if (ucl.show) {
        p = p + geom_line(aes(y = ucl), col = "gray", size = 1)
    }
    
    if (lcl.show) {
        p = p + geom_line(aes(y = lcl), col = "gray", size = 1)
    }
    
    if (band.show) {
        p = p + 
            geom_ribbon(aes(ymin = mean + sigma, 
                            ymax = mean + 2*sigma), alpha = 0.1) +
            geom_ribbon(aes(ymin = pmax(lcl.min, mean - 2*sigma),  
                            ymax = mean - sigma), alpha = 0.1)
        
        orange.sigma = unique(c(
            warn.points(point > mean + sigma, 4, 5),
            warn.points(point < mean - sigma, 4, 5),
            warn.points(point > mean + 2*sigma, 2, 3),
            warn.points(point < mean - 2*sigma, 2, 3)
        ))
    }

    df$warn = "blue"
    if (rule.show) {
        shift.n = round(log(sum(point!=mean), 2) + 3)
        orange = unique(c(orange.sigma,
        warn.points(point > mean - sigma & point < mean + sigma, 15, 15),
        warn.points(point > mean, shift.n, shift.n),
        warn.points(point < mean, shift.n, shift.n)))
        df$warn[orange] = "orange"
    }
    df$warn[point > df$ucl | point < df$lcl] = "red"
    
    p + 
        geom_line(aes(y = point), col = "royalblue3") +
        geom_point(data = df, aes(x = subgroup, y = point, col = warn)) +
        scale_color_manual(values = c("blue" = "royalblue3", "orange" = "orangered", "red" = "red3"), guide = FALSE) +
        labs(x = label.x, y = label.y) +
        theme_bw()
}
```
