--- 
title: "Statistical Process Control in Healthcare with R"
author: "Dwight Barry, Brendan Bettinger, and Andrew Cooper"
date: "`r format(Sys.Date(), '%B %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
github-repo: Rmadillo/spc_healthcare
description: "Using SPC methods in healthcare can be tricky. We show you how to do it correctly, using R."
---

```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = NA)

# Load libraries
library(dplyr)
library(scales)
library(lubridate)
library(forecast) 
library(ggseas)
library(qicharts)
library(bookdown)
library(knitr)
library(ggplot2)
library(ggExtra)
library(gridExtra)
```

# Overview  {#overview}

Statistical process control (SPC) was a triumph of manufacturing analytics, and its success spread across a variety of industries---most improbably, into healthcare. 

Healthcare is rarely compatible with the idea of an assembly line, but lean manufacturing thinking ("Lean") has taken over healthcare management around the world, and SPC methods are common tools in Lean. Unlike in manufacturing, stability is an inherently tricky concept in healthcare, so this has led to much *misuse* of these methods. Bad methods lead to bad inferences, and bad inferences can lead to poor decisions. This book aims to help analysts apply SPC methods more accurately in healthcare, using the statistical software R.  

## If you've never used R

Some BI analysts are apprehensive about getting into R, but if you've ever written a line of SQL or created a formula in an Excel cell, this is no different in concept. Yes, the R language is full of idiosyncracies and outright annoyances, but when you need to accomplish particular goals, it can be fairly easy.   


For example, you can create a *u*-chart with only three lines of code, start to finish:  

<br>  

```{r intro_example, fig.height=3, fig.width=6}
# Load the qicharts package, a simple interface to SPC charts
library(qicharts)

# Load some example data from another R package as an example
data(pcmanufact, package = "qcc")

# Create the u-chart
qic(y = pcmanufact$x, n = pcmanufact$size, chart = "u", main = "Easy u-chart")
```

<br>  

You can find help using `?`, followed by the function name, e.g.,  

```{r help, eval=FALSE}
?qic
```

- You can download R from [https://cran.r-project.org/](https://cran.r-project.org/).  

- You can download RStudio from [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/).  


## What you need

This book was created using R version 3.3.2 and RStudio 1.0.44. Code was tested on Mac OS 10.10.5 (aka Yosemite), Ubuntu 14.04 (aka Trusty Tahr), and Windows Server 2008 R2 (aka piece of shit OS that crashes RStudio when you try to use the Save button).    

Open RStudio and install the packages used in this book by copying and pasting this code into the **Console**:  

```{r install_packs, eval=FALSE}
install.packages("forecast", "ggExtra", "ggseas", "gridExtra", "tidyverse", 
                 "qcc", "qicharts", "scales", dependencies = TRUE)
```


## Book repo

You can submit pull requests for any errors or typos at https://github.com/Rmadillo/spc_healthcare_with_r.  


## About

We are all analysts at *Seattle Children's Hospital* in Seattle, Washington, USA.  

\- Dwight Barry is a Lead Data Scientist in *Enterprise Analytics*. Twitter: @healthstatsdude  

\- Andy Cooper is a Lead Data Scientist in *Ambulatory and Medical Specialties*.  

\- Brendan Bettinger is a Senior Analyst in *Infection Prevention*.  


``` {r spccode, echo = FALSE, fig.height = 3.5}

spc.plot = function(subgroup, point, mean, sigma, k = 3,
                     ucl.show = TRUE, lcl.show = TRUE, 
                     band.show = TRUE, rule.show = TRUE,
                     ucl.max = Inf, lcl.min = -Inf,
                     label.x = "Subgroup", label.y = "Value")
{
    # Plots control chart with ggplot
    # 
    # Args:
    #   subgroup: Subgroup definition (for x-axis)
    #   point: Subgroup sample values (for y-axis)
    #   mean: Process mean value (for center line)
    #   sigma: Process variation value (for control limits)
    #   k: Specification for k-sigma limits above and below center line.
    #      Default is 3.
    #   ucl.show: Visible upper control limit? Default is true.
    #   lcl.show: Visible lower control limit? Default is true.
    #   band.show: Visible bands between 1-2 sigma limits?  Default is true.
    #   rule.show: Highlight run rule indicators in orange?  Default is true.
    #   ucl.max: Maximum feasible value for upper control limit.
    #   lcl.min: Minimum feasible value for lower control limit.
    #   label.x: Specify x-axis label.
    #   label.y: Specify y-axis label.
    
    
    df = data.frame(subgroup, point)
    df$ucl = pmin(ucl.max, mean + k*sigma)
    df$lcl = pmax(lcl.min, mean - k*sigma)
    
    warn.points = function(rule, num, den) {
        sets = mapply(seq, 1:(length(subgroup) - (den - 1)), 
                       den:length(subgroup))
        hits = apply(sets, 2, function(x) sum(rule[x])) >= num
        intersect(c(sets[,hits]), which(rule))
    }
    orange.sigma = numeric()
    
    p = ggplot(data = df, aes(x = subgroup)) +
        geom_hline(yintercept = mean, col = "gray", size = 1)
    
    if (ucl.show) {
        p = p + geom_line(aes(y = ucl), col = "gray", size = 1)
    }
    
    if (lcl.show) {
        p = p + geom_line(aes(y = lcl), col = "gray", size = 1)
    }
    
    if (band.show) {
        p = p + 
            geom_ribbon(aes(ymin = mean + sigma, 
                            ymax = mean + 2*sigma), alpha = 0.1) +
            geom_ribbon(aes(ymin = pmax(lcl.min, mean - 2*sigma),  
                            ymax = mean - sigma), alpha = 0.1)
        
        orange.sigma = unique(c(
            warn.points(point > mean + sigma, 4, 5),
            warn.points(point < mean - sigma, 4, 5),
            warn.points(point > mean + 2*sigma, 2, 3),
            warn.points(point < mean - 2*sigma, 2, 3)
        ))
    }

    df$warn = "blue"
    if (rule.show) {
        shift.n = round(log(sum(point!=mean), 2) + 3)
        orange = unique(c(orange.sigma,
        warn.points(point > mean - sigma & point < mean + sigma, 15, 15),
        warn.points(point > mean, shift.n, shift.n),
        warn.points(point < mean, shift.n, shift.n)))
        df$warn[orange] = "orange"
    }
    df$warn[point > df$ucl | point < df$lcl] = "red"
    
    p + 
        geom_line(aes(y = point), col = "royalblue3") +
        geom_point(data = df, aes(x = subgroup, y = point, col = warn)) +
        scale_color_manual(values = c("blue" = "royalblue3", "orange" = "orangered", "red" = "red3"), guide = FALSE) +
        labs(x = label.x, y = label.y) +
        theme_bw()
}
```
